A Data-Driven Framework for Neural Field Modeling

### 2011-01-12 ltx/first_order_framework/response_to_reviews.tex

\begin{enumerate}
    \item The handling of the (to borrow a term from numerical weather forecasting) `sub-grid scale physics' is inadequate for nervous systems (see, e.g., equation 8 [The nonlinear IntegroDifference Equation]). I do very much appreciate the spectral estimation of spatial and temporal scales. But the model errors are lumped into an iid model error term that is not realistic. For instance, the flow of potassium in the extracellular spaces has a known dynamic that will affect network excitability, generating important sub-grid dynamics that is anything but iid. See, for instance, the example in Ullah et al PLoS Comp Biol 2010 where without such dynamics, state tracking and parameter estimation for a neural system fails. A beautiful overview of sub-grid effects in data assimilation is found in Kalnay's 2003 book. I would very much value the authors comments on these issues. I guess were I to respond to this question, that I might consider extending the model error term to an autoregressive model. But I will let the authors answer this question.

	\emph{To address this point, we have added an extra section to the discussion, `Modeling Assumptions'. Within this section we discuss model mismatch and what we think is a good way to deal with it.}

\item Although we all know the 1970s version of Wilson and Cowan's field equations, there was an intriguing stochasticization of these equations by Benayoun et al PloS Comp Biol 2010. I and others will value the authors' thoughts of how a more palatable stochasticization of these field equations might influence the future development of the data-assimilation framework presented in this paper.

\dean{\emph{We have not added anything to the The authors do not have much to add to this comment. We feel that adding the disturbance to the firing rate is appropriate for this model, which is what we have done, and it agrees with the work by Benayoun et al PloS Comp Biol 2010. By the central limit theorem, we expect the disturbance to be Gaussian and due to the structural organization of the cortex, we expect the disturance to be spatially colored. In future work we plan to add to the framework and show how the disturbance properties can also be estimated from data.}}

### 2011-01-12 ltx/first_order_framework/LaTeX_Manuscript.tex
changes to appendix A and some changes for ken comments

\dean{(Reviewer 1, additional comment d):) Data-driven neural modeling was extended to the continuum approximation by \citet{Galka2008}, where they proposed an estimation framework based on a linear damped wave equation. Using a Kalman filter maximum likelihood framework, they improved on the low resolution electromagnetic tomography (LORETA) method for solving the inverse EEG problem. Another continuum neural field model-based estimator was developed by \citet{Daunizeau2009}, as an extension to the DCM framework. They argue that using a spatially extended continuum field models is superior for explaining cortical function than the neural mass DCM, since field models can explain a richer repertoire of dynamics such as traveling waves and bump solutions.}

@@

The firing rate of the presynaptic neurons is related to the post-synaptic membrane potential by the sigmoidal activation function 
\begin{equation}
	\label{ActivationFunction} f\left( v\left( \mathbf{r}', t \right) \right) = \frac{1}{1 + \exp \left( \varsigma \left( v_0 - v\left(\mathbf{r}',t\right) \right) \right)}. 
\end{equation}
The parameter $v_0$ describes the firing threshold of the neural populations and $\varsigma$ governs the slope of the sigmoid. By substituting equation~\ref{RateBasedInteractions} into equation~\ref{SpikesToPotential} we get the spatiotemporal model 
\begin{equation}
	\label{FullDoubleIntModel} v\left(\mathbf{r},t\right) =
	\int_{-\infty}^t 
	h\left(t - t'\right) \int_\Omega
	w\left(\mathbf{r},\mathbf{r}'\right) 
	f\left( v\left( \mathbf{r}',t' \right)\right)
	\, d\mathbf{r}'dt'.
\end{equation}
\dean{(reviewer 2, comment 3:) To arrive at the standard integro-differential equation form of the model, we use the fact that the synaptic response kernel is a Green's function of a linear differential equation defined by the differential operator $\textrm{D}=d/dt + \zeta$. A Green's function satisfies
\begin{equation}
	\label{GreensFuncDef} \textrm{D}h\left( t \right) = \delta \left( t \right), 
\end{equation} 
where $\delta(t)$ is the Dirac-delta function. Applying the differential operator $\textrm{D}$ to equation~\ref{FullDoubleIntModel} gives
\begin{align}
 \textrm{D}v\left(\mathbf r,t\right)&= \textrm{D}\left(h\ast g\right)\left(\mathbf r,t\right)\\
&=\left(\textrm{D}h\ast g\right)\left(\mathbf r,t\right)\\
&=\left(\delta \left(t\right)\ast g\right)\left(\mathbf r,t\right)\\
&=g\left(\mathbf r,t\right)
\end{align}
where $\ast$ denotes the convolution operator.} This gives the standard form of the model
\begin{equation}
	\label{FinalFormContinuous} 
	\frac{dv\left( \mathbf{r},t \right)}{dt} + \zeta v\left( \mathbf{r},t \right) = \int_\Omega {w\left( \mathbf{r},\mathbf{r}' \right)f\left( {v\left( \mathbf{r}',t \right)} \right)\, d\mathbf{r}'}. 
\end{equation}
To arrive at the integro-difference equation (IDE) form of the model, we discretize time using a first-order Euler method (see \ref{App:Time Discretization} for derivation) giving 
\begin{equation}
	\label{DiscreteTimeModel} 
	v_{t+T_s}\left(\mathbf{r}\right) = 
	\xi v_t\left(\mathbf{r}\right) + 
	T_s \int_\Omega { 
	    w\left(\mathbf{r},\mathbf{r}'\right)
	    f\left(v_t\left(\mathbf{r}'\right)\right) 
	\, d\mathbf{r}'} 
	+ e_t\left(\mathbf{r}\right), 
\end{equation}
where $T_s$ is the time step, $\xi = 1-T_s\zeta $ and $e_t(\mathbf{r})$ is an $i.i.d.$ disturbance such that $e_t(\mathbf{r})\sim\mathcal{GP}(\mathbf 0,\gamma(\mathbf{r}-\mathbf{r}'))$. Here $\mathcal{GP}(\mathbf 0,\gamma(\mathbf{r}-\mathbf{r}'))$ denotes a zero mean Gaussian process with spatial covariance function $\gamma(\mathbf{r}-\mathbf{r}')$ (\cite{Rasmussen2005}). The disturbance is added to account for model uncertainty and unmodeled inputs. To simplify the notation, the index of the future time sample, $t+T_s$, shall be referred to as $t+1$ throughout the rest of the paper.

@@

\parham{(Reviewer1, Additional comments 2:) The preservation of the system's stability, identifiability, observability, and controllability properties is not guaranteed with the discretization. Observability is assumed and then verified with the estimation results. Here we show that if $0 < T_s < \tau$, where the synaptic time constant $\tau = 1/\zeta$,  the discretized IDE is stable. To do this, the model is written as the series
\begin{align}
	v_1(\mathbf{r}) &= \xi v_0(\mathbf{r}) + T_s g_0(\mathbf{r}) \\
	v_2(\mathbf{r})	&= \xi^2 v_0(\mathbf{r}) + T_s \left(\xi g_0(\mathbf{r}) + g_1(\mathbf{r}) \right) \\
	\vdots \, \, \, \, \, &= \, \, \, \, \, \vdots \nonumber \\
	v_{t+1}(\mathbf r) &= \xi^{t+1} v_0\left(\mathbf{r} \right) + T_s \sum_{i=0}^t \xi^i g_{t-i}(\mathbf{r}), \label{eq:fieldmodelseries}
\end{align}
where
\begin{equation}
	g_{t}(\mathbf{r}) = \int_\Omega { w\left(\mathbf{r}-\mathbf{r}'\right) f\left(v_t\left(\mathbf{r}'\right)\right) d\mathbf{r}'}.
\end{equation}
Now, $g_t(\mathbf r)$ is always bounded since $0 \leq f(v_t(\mathbf r)) \leq 1$ and the area under the kernel is finite. To assess the stability of the discrete time system we assume that $g_t(\mathbf{r})$ is maximally saturated over the entire field for all time, which we denote as $g_{max}$. Substituting this into equation~\ref{eq:fieldmodelseries} yields
\begin{align}\label{vrecursiveinlimit}
	v_{t+1}(\mathbf r) &= \xi^{t+1}v_0(\mathbf r)+ g_{max} T_s \sum_{i=0}^t \xi^i \nonumber \\
	&=\xi^{t+1}v_0(\mathbf r) + \frac{ g_{max} T_s(1-\xi^{t+1})}{1-\xi}.
\end{align}
To study the asymptotic behaviour of the system we consider the limit as $t$ approaches infinity. Noting $0 < \xi <1$  we use a property of the geometric series to simplify equation \eqref{vrecursiveinlimit} to
\begin{equation}\label{simplifiedvrecursiveinlimit}
	v_{t+1}(\mathbf r)=\xi^{t+1}v_0(\mathbf r)+\frac{g_{max} T_s}{1-\xi}.
\end{equation}
Substituting $\xi=1-\zeta T_s$ into equation \eqref{simplifiedvrecursiveinlimit} gives
\begin{equation}
	v_{t+1}(\mathbf r)=(1-\zeta T_s)^{t+1}v_0(\mathbf r)+\frac{g_{max}}{\zeta}.
\end{equation}
From this equation it can be seen that if $0 < T_s < \tau$, the system is stable. The sampling period was set to $T_s = 0.001 s$ and the time constant was set to $\tau = 0.01 s^{-1}$. Therefore, the discretization preserved the stability properties of the system.}

@@

\section{Equivalence to Galerkin's Projection}\label{App:Galerkin}
\dean{To show how equation~\ref{eq:decompRHS} is equivalent to the Galerkin projection, we define the residual between the field and the decomposition as 
\begin{align}\label{eq:BasisFunctionResidual}
	\Xi(\mathbf{x}_t,v_{t+1}&(\mathbf{r})) = v_{t+1}(\mathbf{r}) \nonumber \\
	&- T_s\int_\Omega{f(\boldsymbol{\phi}^{\top}(\mathbf{r}')\mathbf{x}_t )\boldsymbol{\psi}^{\top}(\mathbf{r}-\mathbf{r}') \, d\mathbf{r}'}\boldsymbol{\theta} \nonumber \\
	&+ \xi\boldsymbol{\phi}^{\top}(\mathbf{r})\mathbf{x}_t + \mathbf{e}_t(\mathbf{r}).
\end{align} 
The Galerkin projection finds $\mathbf{x}_t$, such that the inner product of the residual with all of the basis functions satisfies
\begin{equation}\label{eq:GalerkinProjection}
	G_i = \int_{\Omega}\Xi(\mathbf{x}_t,v_{t+1}(\mathbf{r})) \phi_i(\mathbf{r}) d\mathbf{r} = 0,
\end{equation}
where $i=0,1,\hdots,n_x-1,$ and $n_x$ is the number of states and basis functions. If equation~\ref{eq:GalerkinProjection} is satisfied the basis functions are orthogonal to the residual. By substituting equation~\ref{eq:BasisFunctionResidual} into equation~\ref{eq:GalerkinProjection} and rearranging we get equation~\ref{eq:decompRHS} of the main text. Note, if Galerkin's method was used estimate the states, $n_x$ coupled ODEs would need to be solved independently (each $G_i$) to estimate the state vector, which is a standard method in fluid dynamics and economics modeling. The method in this paper takes a different approach by forming the state-space model, allowing for more efficient estimation. }

>>>>

\dean{(Reviewer 1, additional comment d):) Data-driven neural modeling was extended to the continuum approximation by \citet{Galka2008}, where they proposed an estimation framework based on a linear damped wave equation. Using a Kalman filter maximum likelihood framework, they improved on the low resolution electromagnetic tomography (LORETA) method for solving the inverse EEG problem. Another continuum neural field model-based estimator was developed by \citet{Daunizeau2009}, as an extension to the DCM framework. \parham{They argue that using a spatially extended continuum field model is superior for explaining cortical function than the neural mass DCM, since field models can explain a richer repertoire of dynamics such as traveling waves and bump solutions.}}

@@

The firing rate of the presynaptic neurons is related to the post-synaptic membrane potential by the sigmoidal activation function 
\begin{equation}
	\label{ActivationFunction} f\left( v\left( \mathbf{r}', t \right) \right) = \frac{1}{1 + \exp \left( \varsigma \left( v_0 - v\left(\mathbf{r}',t\right) \right) \right)}. 
\end{equation}
The parameter $v_0$ describes the firing threshold of the neural populations and $\varsigma$ governs the slope of the sigmoid. By substituting equation~\ref{RateBasedInteractions} into equation~\ref{SpikesToPotential} we get the spatiotemporal model 
\begin{equation}
	\label{FullDoubleIntModel} v\left(\mathbf{r},t\right) =
	\int_{-\infty}^t 
	h\left(t - t'\right) \int_\Omega
	w\left(\mathbf{r},\mathbf{r}'\right) 
	f\left( v\left( \mathbf{r}',t' \right)\right)
	\, d\mathbf{r}'dt'.
\end{equation}
\parham{The post-synaptic membrane potential is in fact the temporal convolution between the synaptic kernel and the average firing rate
\begin{equation}\label{FullConvolutionModel}
 v\left(\mathbf r,t\right)= \left(h\ast g\right)\left(\mathbf r,t\right)
\end{equation}
where $\ast$ denotes the convolution operator.}
\dean{(reviewer 2, comment 3:) \parham{To obtain (To arrive at)} the standard integro-differential equation form of the model, we use the fact that the synaptic response kernel is a Green's function of a linear differential equation defined by the differential operator $\textrm{D}=d/dt + \zeta$. A Green's function satisfies
\begin{equation}
	\label{GreensFuncDef} \textrm{D}h\left( t \right) = \delta \left( t \right), 
\end{equation} 
where $\delta(t)$ is the Dirac-delta function. Applying the differential operator $\textrm{D}$ \parham{to equation~\ref{FullConvolutionModel}} gives
\begin{align}
 \textrm{D}v\left(\mathbf r,t\right)&= \textrm{D}\left(h\ast g\right)\left(\mathbf r,t\right)\\
&=\left(\textrm{D}h\ast g\right)\left(\mathbf r,t\right)\\
&=\left(\delta \left(t\right)\ast g\right)\left(\mathbf r,t\right)\\
&=g\left(\mathbf r,t\right)
\end{align}
} This gives the standard form of the model
\begin{equation}
	\label{FinalFormContinuous} 
	\frac{dv\left( \mathbf{r},t \right)}{dt} + \zeta v\left( \mathbf{r},t \right) = \int_\Omega {w\left( \mathbf{r},\mathbf{r}' \right)f\left( {v\left( \mathbf{r}',t \right)} \right)\, d\mathbf{r}'}. 
\end{equation}
To arrive at the integro-difference equation (IDE) form of the model, we discretize time using a first-order Euler method (see \ref{App:Time Discretization} for derivation) giving 
\begin{equation}
	\label{DiscreteTimeModel} 
	v_{t+T_s}\left(\mathbf{r}\right) = 
	\xi v_t\left(\mathbf{r}\right) + 
	T_s \int_\Omega { 
	    w\left(\mathbf{r},\mathbf{r}'\right)
	    f\left(v_t\left(\mathbf{r}'\right)\right) 
	\, d\mathbf{r}'} 
	+ e_t\left(\mathbf{r}\right), 
\end{equation}
where $T_s$ is the time step, $\xi = 1-T_s\zeta $ and $e_t(\mathbf{r})$ is an $i.i.d.$ disturbance such that $e_t(\mathbf{r})\sim\mathcal{GP}(\mathbf 0,\gamma(\mathbf{r}-\mathbf{r}'))$. Here $\mathcal{GP}(\mathbf 0,\gamma(\mathbf{r}-\mathbf{r}'))$ denotes a zero mean Gaussian process with spatial covariance function $\gamma(\mathbf{r}-\mathbf{r}')$ (\cite{Rasmussen2005}). The disturbance is added to account for model uncertainty and unmodeled inputs. To simplify the notation, the index of the future time sample, $t+T_s$, shall be referred to as $t+1$ throughout the rest of the paper.

@@

\parham{(Reviewer1, Additional comments 2:) The analysis of the error introduced by Euler discretization method can be found in \citet{Butcher2008}. Here we only study the stability of the discretized model, we show that if $0 < T_s < \tau$, where the synaptic time constant $\tau = 1/\zeta$,  the discretized IDE is stable. To do this, the model is written as the series
\begin{align}
	v_1(\mathbf{r}) &= \xi v_0(\mathbf{r}) + T_s g_0(\mathbf{r}) \\
	v_2(\mathbf{r})	&= \xi^2 v_0(\mathbf{r}) + T_s \left(\xi g_0(\mathbf{r}) + g_1(\mathbf{r}) \right) \\
	\vdots \, \, \, \, \, &= \, \, \, \, \, \vdots \nonumber \\
	v_{t+1}(\mathbf r) &= \xi^{t+1} v_0\left(\mathbf{r} \right) + T_s \sum_{i=0}^t \xi^i g_{t-i}(\mathbf{r}), \label{eq:fieldmodelseries}
\end{align}
where
\begin{equation}
	g_{t}(\mathbf{r}) = \int_\Omega { w\left(\mathbf{r}-\mathbf{r}'\right) f\left(v_t\left(\mathbf{r}'\right)\right) d\mathbf{r}'}.
\end{equation}
Since  $0 \leq f(v_t(\mathbf r)) \leq 1$ and the kernel is semi-compactly supported, $g_t(\mathbf r)$ is always bounded . To assess the stability of the discrete time system we assume that $g_t(\mathbf{r})$ is maximally saturated over the entire field for all time, which we denote as $g_{max}$. Substituting this into equation~\ref{eq:fieldmodelseries} yields
\begin{align}\label{vrecursiveinlimit}
	v_{t+1}(\mathbf r) &= \xi^{t+1}v_0(\mathbf r)+ g_{max} T_s \sum_{i=0}^t \xi^i \nonumber \\
	&=\xi^{t+1}v_0(\mathbf r) + \frac{ g_{max} T_s(1-\xi^{t+1})}{1-\xi}.
\end{align}
To study the asymptotic behaviour of the system we consider the limit as $t$ approaches infinity. Noting $0 < \xi <1$  we use a property of the geometric series to simplify equation \eqref{vrecursiveinlimit} to
\begin{equation}\label{simplifiedvrecursiveinlimit}
	v_{t+1}(\mathbf r)=\xi^{t+1}v_0(\mathbf r)+\frac{g_{max} T_s}{1-\xi}.
\end{equation}
Substituting $\xi=1-\zeta T_s$ into equation \eqref{simplifiedvrecursiveinlimit} gives
\begin{equation}
	v_{t+1}(\mathbf r)=(1-\zeta T_s)^{t+1}v_0(\mathbf r)+\frac{g_{max}}{\zeta}.
\end{equation}
From this equation it can be seen that if $0 < T_s < \tau$, the system is stable. The sampling period was set to $T_s = 0.001 s$ and the time constant was set to $\tau = 0.01 s^{-1}$. Therefore, the discretization preserved the stability properties of the system.}

@@

\section{Equivalence to Galerkin's Projection}\label{App:Galerkin}
\dean{To show how equation~\ref{eq:decompRHS} is equivalent to the Galerkin projection, we define the residual between the field and the decomposition as 
\begin{align}\label{eq:BasisFunctionResidual}
	\Xi(\mathbf{x}_t,v_{t+1}&(\mathbf{r})) = v_{t+1}(\mathbf{r}) \nonumber \\
	&- T_s\int_\Omega{f(\boldsymbol{\phi}^{\top}(\mathbf{r}')\mathbf{x}_t )\boldsymbol{\psi}^{\top}(\mathbf{r}-\mathbf{r}') \, d\mathbf{r}'}\boldsymbol{\theta} \nonumber \\
	&+ \xi\boldsymbol{\phi}^{\top}(\mathbf{r})\mathbf{x}_t + \mathbf{e}_t(\mathbf{r}).
\end{align} 
The Galerkin projection finds $\mathbf{x}_t$, such that the inner product of the residual with all of the basis functions satisfies
\begin{equation}\label{eq:GalerkinProjection}
	G_i = \int_{\Omega}\Xi(\mathbf{x}_t,v_{t+1}(\mathbf{r})) \phi_i(\mathbf{r}) d\mathbf{r} = 0,
\end{equation}
where $i=0,1,\hdots,n_x-1,$ and $n_x$ is the number of states and basis functions. If equation~\ref{eq:GalerkinProjection} is satisfied the basis functions are orthogonal to the residual. By substituting equation~\ref{eq:BasisFunctionResidual} into equation~\ref{eq:GalerkinProjection} and rearranging we get equation~\ref{eq:decompRHS} of the main text. Note, if Galerkin's method was used to estimate the states, $n_x$ coupled ODEs would need to be solved independently (each $G_i$) to estimate the state vector, which is a standard method in fluid dynamics and economics modeling. The method in this paper takes a different approach by forming the state-space model, allowing for more efficient estimation. }
